---
title: "FinalReport"
author: "Nick Mandarano and Patrick McHugh"
date: "4/26/2022"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


# Loading Libraries
library(tidyverse)
library(leaps)
library(MASS)
library(ggplot2)
library(gridExtra)
library(car)
library(dplyr)
library(readr)
library(corrplot)
```

## Introduction

## Variable Selection

Our data comes from four different datasets. We used three of Riguang Wen's [datasets](https://figshare.com/articles/dataset/NBA_data/5414170) from figshare.com -- \texttt{players cv}, \texttt{players salary}, and \texttt{players stat}. We also used a [dataset](https://zenodo.org/record/3750329#.YkT6YW7MJAe) called \texttt{NBA RS 2020-1950 Stats} uploaded to zenodo.org by Pablo Gomez and Sandra Giral. From these datasets, we considered the following variables.

| Variable      | Description               | Type          | Source                  |
|:--------------|:-------------------------:|:-------------:|:-----------------------:|
| Player        | Name of player            | Character     | \texttt{players stat}   |
| Age           | Age of player             | Numeric       | \texttt{players stat}   |
| G             | Games played              | Numeric       | \texttt{players stat}   |
| GS            | Games started             | Numeric       | \texttt{players stat}   |
| MP            | Minutes played            | Numeric       | \texttt{players stat}   |
| PER           | Player efficiency rating  | Numeric       | \texttt{players stat}   |
| PTS           | Points                    | Numeric       | \texttt{NBA RS 2020-1950 Stats} |
| X3PAr         | 3PA/FGA                   | Numeric       | \texttt{players stat}   |
| FTr           | FTA/FGA                   | Numeric       | \texttt{players stat}   |
| TS            | True shooting percentage  | Numeric       | \texttt{NBA RS 2020-1950 Stats} |
| ORB           | Offensive rebounds        | Numeric       | \texttt{players stat}   |
| DRB           | Defensive rebounds        | Numeric       | \texttt{players stat}   |
| TRB           | Total rebounds            | Numeric       | \texttt{players stat}   |
| AST           | Assists                   | Numeric       | \texttt{players stat}   |
| STL           | Steals                    | Numeric       | \texttt{players stat}   |
| BLK           | Blocks                    | Numeric       | \texttt{players stat}   |
| TOV           | Turnovers                 | Numeric       | \texttt{players stat}   |
| USG           | Usage percentage          | Numeric       | \texttt{players stat}   |
| ORtg          | Offensive rating          | Numeric       | \texttt{players stat}   |
| DRtg          | Defensive rating          | Numeric       | \texttt{players stat}   |
| OWS           | Offensive win shares      | Numeric       | \texttt{players stat}   |
| DWS           | Defensive win shares      | Numeric       | \texttt{players stat}   |
| WS            | Win shares                | Numeric       | \texttt{players stat}   |
| WS.48         | Win shares per 48 minutes | Numeric       | \texttt{players stat}   |
| OBPM          | Offensive box +/-         | Numeric       | \texttt{players stat}   |
| DBPM          | Defensive box +/-         | Numeric       | \texttt{players stat}   |
| BPM           | Box +/-                   | Numeric       | \texttt{players stat}   |
| VORP          | Value over replacement player | Numeric   | \texttt{players stat}   |
| Pos           | Position                  | Factor        | \texttt{players salary} |
| Ht            | Height in inches          | Numeric       | \texttt{players salary} |
| Wt            | Weight in pounds          | Numeric       | \texttt{players salary} |
| PwrSix        | Power Six College?        | Indicator     | \texttt{players cv}     |
| International | International Player?     | Indicator     | \texttt{players cv}     |
| Salary        | Salary in dollars         | Numeric       | \texttt{players salary} |

### Exploratory Data Analysis

Immediately we can recognize that some variables are functions of others and therefore do not need to be considered. Specifically, \texttt{BPM = OBPM + DBPM}, so there is no need to include \texttt{BPM} in our model. Similarly, \texttt{WS = OWS + DWS} and \texttt{TRB = ORB + DRB}, so we can exclude \texttt{WS} and \texttt{TRB} from consideration if we include \texttt{OWS}, \texttt{DWS}, \texttt{ORB} and \texttt{DRB} in our model. 

From our EDA, we concluded that a log transformation of the response variable, salary, would be appropriate based on the skewed distribution.

We also noticed that many of the covariates had skewed distributions, or did not have linear marginal relationships with the response log(salary). We attempted to make these variables approximately normally distirbutied, with an approximately linear marginal relationship with the response. We experimented with log, square root, and squaring transformations, and ended up considering the following transformations in our model:

* VORP: We min/max normalized this between 0 and 1 to eliminate negative values, and then took the log
* OWS: We min/max normalized this between 0 and 1 to eliminate negative values, and then took the log
* DWS: Square root
* PTS: Square root
* FTr: Square root
* BLK: log
* STL: log
* GS: log
* ORB: log

Below we used VORP as an example; we can see the distribution of VORP and its plot against log(salary) pre and post transformation:



```{r}
data <- read.csv("data/PlayerData.csv")[,-1] %>% mutate(logsal = log(SALARY))
attach(data)
Pos <- as.factor(Pos)
International <- as.factor(International)
Conference <- as.factor(Conference)
Multiteam <- as.factor(Multiteam)
Pos_cat <- as.factor(Pos_cat)

vorp_hist = data %>% ggplot(aes(x=VORP)) + geom_histogram() + ggtitle("Histogram of VORP")
vorp_scatter = data %>% ggplot(aes(x=VORP)) + geom_point(aes(y=logsal)) + ggtitle("Scatterplot of VORP vs. log salary")
adj_vorp_hist = data %>% ggplot(aes(x=vorp_adj)) + geom_histogram() + ggtitle("Histogram of transformed VORP")
adj_vorp_scatter = data %>% ggplot(aes(x=vorp_adj)) + geom_point(aes(y=logsal)) + ggtitle("Scatterplot of transformed VORP vs. log salary")
grid.arrange(vorp_hist, vorp_scatter, adj_vorp_hist, adj_vorp_scatter, nrow = 2)
```

After the EDA

### Correlation Matrix
```{r}
corrplot(cor(data[sapply(data, is.numeric)]))
cor(PTS, MP)
cor(WS.48, PER)
cor(DBPM,DRtg)
cor(ORtg, TS.)
# Which did we remove from here and why?
```

### Stepwise Regression Models
```{r}

null = lm(logsal ~ 1, data = data)
full = lm(logsal ~ ., data = data)
summary(full)

bic = log(nrow(data))
stepbic_model <- stepAIC(object = null, scope = list(lower = null, upper = full),
direction = "both", k = bic)
stepaic_model <- stepAIC(object = null, scope = list(lower = null, upper = full),
direction = "both", k = 2)
```

### Variance Inflation Factors
```{r}
full = lm(logsal ~ ., data = data)
summary(full)
#vif(full)[,3]
#full2 = lm(logsal ~ . - WS.48 - PER - ORtg - PTS, data = data) # Sequentially removed highest VIF until next iteration lowered both Adj R2 and Mult R2
```

### Added Variable Plots

## Diagnostics

### Leverage Points

### Cook's Distances

### Outliers

## Model Fitting

### Test for Interaction

### Test for Higher Order

### Model Validation

### Final Model

## Discussion

## Conclusion